A Certified Interpreter for the List Machine Benchmark

### 2020-06-08 papers/SBLP2020/paper.tex
Starting the section about the benchmark

@@

@@

@@

@@

@@

Another useful data type is the finite type, \ensuremath{\D{Fin}}\footnote{Note that Agda supports the overloading of data type constructor names.
Constructor \ensuremath{\Con{zero}} can refer to type \ensuremath{\D{\mathbb{N}}} or \ensuremath{\D{Fin}}, depending on the
context where the name is used.}, which is defined in
Agda's standard library as:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathkw{data}\;\D{Fin}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{zero}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;(\Con{suc}\;\V{n}){}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{suc}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{Fin}\;(\Con{suc}\;\V{n}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Type \ensuremath{\D{Fin}\;\V{n}} has exactly \ensuremath{\V{n}} inhabitants
(elements), i.e., it is isomorphic to the set $\{0,...,n - 1\}$.
An application of such type is to define a safe vector
lookup function, which avoids the access of invalid positions.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\F{lookup}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{A}\;\V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{Vec}\;\V{A}\;\V{n}\;\to \;\V{A}{}\<[E]%
\\
\>[3]{}\F{lookup}\;\Con{zero}\;(\V{x}\;\Con{::}\;\anonymous )\;\mathrel{=}\;\V{x}{}\<[E]%
\\
\>[3]{}\F{lookup}\;(\Con{suc}\;\V{idx})\;(\anonymous \;\Con{::}\;\V{xs})\;\mathrel{=}\;\F{lookup}\;\V{idx}\;\V{xs}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Thanks to the propositions-as-types principle,\footnote{It is also known as
  Curry-Howard ``isomorphism''~\cite{Sorensen2006}.} we can interpret
types as logical formulas and terms as proofs. An example is the
representation of equality as the following Agda type:

@@

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%
%% Title and authors %%
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%
%% Abstract          %%
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%
%% ACM codes for areas %%
%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction      %%
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%
%% Overview of Agda  %%
%%%%%%%%%%%%%%%%%%%%%%%

@@

Another useful data type is the finite type,
\ensuremath{\D{Fin}}\footnote{Note that Agda supports the overloading of
data type constructor names. Constructor \ensuremath{\Con{zero}} can refer to
type \ensuremath{\D{\mathbb{N}}} or \ensuremath{\D{Fin}}, depending on the context where the name
is used.}, which is defined in Agda's standard library as:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathkw{data}\;\D{Fin}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{zero}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;(\Con{suc}\;\V{n}){}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{suc}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{Fin}\;(\Con{suc}\;\V{n}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Type \ensuremath{\D{Fin}\;\V{n}} has exactly \ensuremath{\V{n}} inhabitants
(elements), i.e., it is isomorphic to the set $\{0,...,n - 1\}$.
An application of such type is to define a safe vector
lookup function, which avoids the access of invalid positions.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\F{lookup}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{A}\;\V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{Vec}\;\V{A}\;\V{n}\;\to \;\V{A}{}\<[E]%
\\
\>[3]{}\F{lookup}\;\Con{zero}\;(\V{x}\;\Con{::}\;\anonymous )\;\mathrel{=}\;\V{x}{}\<[E]%
\\
\>[3]{}\F{lookup}\;(\Con{suc}\;\V{idx})\;(\anonymous \;\Con{::}\;\V{xs})\;\mathrel{=}\;\F{lookup}\;\V{idx}\;\V{xs}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Thanks to the propositions-as-types principle,\footnote{It is also known as
  Curry-Howard ``isomorphism''~\cite{Sorensen2006}.} we can interpret
types as logical formulas and terms as proofs. An example is the
representation of equality as the following Agda type:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% List machine benchmark %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The list machine is a simple pointer virtual machine whose values
are empty lists and cons-cells:
\[
a ::= nil\,\mid\,cons(a_1,a_2)
\]
We let meta-variable $a$ to denote an arbitrary value, $v$ denotes a variable
and $l$ a program label. Following common practice, all meta-variables can
appear primed or subscripted. The syntax of the virtual machine instructions are
presented next and their meaning is as usual.
\[
\begin{array}{rcll}
  i & ::=  & \text{jump }l                       & \text{(jump instruction)}\\
    & \mid & \text{branch-if-nil $v$}            & \text{(if $v = nil$ goto $l$)}\\
    & \mid & \text{fetch-field $v$ 0 $v'$}       & \text{(fetch the head of $v$ into $v'$)}\\
    & \mid & \text{fetch-field  $v$ 1 $v'$}      & \text{(fetch the tail of $v$ into $v'$)}\\
    & \mid & \text{cons $v_0$ $v_1$ $v'$}        & \text{(make a cons cell in v')} \\
    & \mid & \text{halt}                         & \text{(finishes execution)}\\
    & \mid & i_1;i_2                             & \text{(sequential composition)}\\
  p & ::=  & l_i \,:\,i\,;\,p                    & \text{(program: sequence of blocks)}\\
    & \mid & \text{end}                          & \text{(end of block list)}\\
\end{array}
\]
Programs are just a sequence of blocks which begin with a unique label.

Each program variable is assigned a list type which is used to guarantee the safety of execute
fetch-field that demands non-empty list arguments and branch instructions refine the argument type
according to which branch is taken. In order to express such refinements, types are subject to a
subtyping relation. We let the meta-variable $\tau$ denote an arbitrary type. 
\[
\begin{array}{rcll}
  \tau & ::=  & \text{nil} & \text{(type for empty lists)}\\
       & \mid & \text{list }\tau & \text{(lists whose elements have type $\tau$)}\\
       & \mid & \text{listcons }\tau & \text{(non-empty lists of $\tau$)}\\
\end{array}
\]
We let notation $\tau \subset \tau'$
denote the subtyping judgement which is defined as follows.
\[
\begin{array}{ccc}
  \inference{}
            {\tau\subset \tau}
            [refl]
  &
  \inference{}
            {nil \subset \tau}
            [nil]
  &
  \inference{\tau \subset \tau'}
            {list\:\tau\subset list\:\tau'}
            [list]\\\\
  \multicolumn{3}{c}{
  \inference{\tau \subset \tau'}
            {listcons\:\tau\subset list\:\tau'}
            [listcons]} \\\\
            \multicolumn{3}{c}{
            \inference{\tau \subset \tau'}
            {listcons\:\tau\subset listcons\:\tau'}
            [mixed]} \\\\
\end{array}
\]
Basically, the subtyping relation specifies that $nil$ (empty list type) is
subtype of any list type and $listcons\:\tau$ is a subtype of the $list\:\tau$.
